---
title: "Chambana Health Dashboard"
format: 
  dashboard:
    theme: flatly
    orientation: rows
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

# Load all packages once
library(sf)
library(leaflet)
library(dplyr)
library(plotly)
library(tidycensus)
library(htmltools)

# Set Census API key
census_api_key("8a4517f7bed332fc21d9b4f2173029e771da2eb2", overwrite = TRUE, install = FALSE)

# Read geographic data once and convert GEOID to character
cu_geo <- st_read("/Users/jakecox/rstudio/cu-geo/cu-tracts-fixed.json", quiet = TRUE)
cu_geo$GEOID <- as.character(cu_geo$GEOID)
```

# Social Determinants of Health

## Row {height="55%"}

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Poverty Rate by Census Tract"

# Read data files
cu_poverty <- read.csv("/Users/jakecox/rstudio/cu-geo/cu-poverty_revised.csv")
cu_sdoh <- read.csv("/Users/jakecox/rstudio/cu-geo/cu-sdoh.csv")

# Join the data
cu_data_sdoh <- cu_geo %>%
  left_join(cu_poverty, by = "NAMELSAD") %>%
  left_join(cu_sdoh, by = "NAMELSAD")

# Create color palette
pal_poverty <- colorNumeric(
  palette = "YlOrRd",
  domain = cu_data_sdoh$Poverty.Rate[!is.na(cu_data_sdoh$Poverty.Rate)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_sdoh) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_poverty(Poverty.Rate),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "Poverty Rate: ", round(Poverty.Rate, 1), "%"
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_poverty,
    values = ~Poverty.Rate,
    title = "Poverty Rate (%)",
    opacity = 0.7,
    na.label = "No data"
  )
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Median Household Income"

# Create color palette
pal_income <- colorNumeric(
  palette = "Greens",
  domain = cu_data_sdoh$median_income[!is.na(cu_data_sdoh$median_income)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_sdoh) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_income(median_income),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "Median Income: $", format(median_income, big.mark = ",")
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_income,
    values = ~median_income,
    title = "Median Income ($)",
    opacity = 0.7,
    na.label = "No data",
    labFormat = labelFormat(prefix = "$", big.mark = ",")
  )
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Health Insurance Coverage"

# Create color palette
pal_insurance <- colorNumeric(
  palette = "Blues",
  domain = cu_data_sdoh$insurance_coverage[!is.na(cu_data_sdoh$insurance_coverage)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_sdoh) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_insurance(insurance_coverage),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "Insurance Coverage: ", round(insurance_coverage, 1), "%"
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_insurance,
    values = ~insurance_coverage,
    title = "Coverage (%)",
    opacity = 0.7,
    na.label = "No data"
  )
```

## Row {height="35%"}

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Poverty Rate Distribution"

plot_ly(
  x = ~cu_data_sdoh$Poverty.Rate,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#fc8d59", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Poverty Rate (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 50, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Income Distribution"

plot_ly(
  x = ~cu_data_sdoh$median_income,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#31a354", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Median Household Income ($)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 50, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Insurance Coverage Distribution"

plot_ly(
  x = ~cu_data_sdoh$insurance_coverage,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#3182bd", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Insurance Coverage (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 50, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {height="10%"}

### Column
```{r}
#| echo: false

div(
  style = "padding: 10px; background-color: #f8f9fa; border-left: 4px solid #6c757d; margin: 5px; font-size: 13px;",
  tags$p(
    style = "margin: 0; color: #495057;",
    tags$strong("Source: "),
    "U.S. Census Bureau, American Community Survey 5-Year Estimates (2018-2022)."
  )
)
```

# Economic & Access Indicators

## Row {height="55%"}

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Educational Attainment (Bachelor's+)"

# Get educational attainment data
education_data <- get_acs(
  geography = "tract",
  state = "IL",
  county = "Champaign",
  variables = c(
    bachelors = "B15003_022",
    masters = "B15003_023",
    professional = "B15003_024",
    doctorate = "B15003_025",
    total_pop = "B15003_001"
  ),
  year = 2022,
  survey = "acs5",
  output = "wide"
)
education_data$pct_bachelor_plus <- ((education_data$bachelorsE + education_data$mastersE + 
                                       education_data$professionalE + education_data$doctorateE) / 
                                       education_data$total_popE) * 100
education_data$GEOID <- as.character(education_data$GEOID)
education_data <- education_data[, c("GEOID", "pct_bachelor_plus")]

# Get unemployment data
unemployment_data <- get_acs(
  geography = "tract",
  state = "IL",
  county = "Champaign",
  variables = c(
    unemployed = "B23025_005",
    labor_force = "B23025_003"
  ),
  year = 2022,
  survey = "acs5",
  output = "wide"
)
unemployment_data$unemployment_rate <- (unemployment_data$unemployedE / unemployment_data$labor_forceE) * 100
unemployment_data$GEOID <- as.character(unemployment_data$GEOID)
unemployment_data <- unemployment_data[, c("GEOID", "unemployment_rate")]

# Get vehicle availability data
vehicle_data <- get_acs(
  geography = "tract",
  state = "IL",
  county = "Champaign",
  variables = c(
    no_vehicle = "B08201_002",
    total_households = "B08201_001"
  ),
  year = 2022,
  survey = "acs5",
  output = "wide"
)
vehicle_data$pct_no_vehicle <- (vehicle_data$no_vehicleE / vehicle_data$total_householdsE) * 100
vehicle_data$GEOID <- as.character(vehicle_data$GEOID)
vehicle_data <- vehicle_data[, c("GEOID", "pct_no_vehicle")]

# Join all data to geography
cu_data_econ <- cu_geo %>%
  left_join(education_data, by = "GEOID") %>%
  left_join(unemployment_data, by = "GEOID") %>%
  left_join(vehicle_data, by = "GEOID")

# Create color palette
pal_education <- colorNumeric(
  palette = "Purples",
  domain = cu_data_econ$pct_bachelor_plus[!is.na(cu_data_econ$pct_bachelor_plus)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_econ) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_education(pct_bachelor_plus),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "Bachelor's+: ", round(pct_bachelor_plus, 1), "%"
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_education,
    values = ~pct_bachelor_plus,
    title = "Bachelor's+ (%)",
    opacity = 0.7,
    na.label = "No data"
  )
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Unemployment Rate"

# Create color palette
pal_unemployment <- colorNumeric(
  palette = "Oranges",
  domain = cu_data_econ$unemployment_rate[!is.na(cu_data_econ$unemployment_rate)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_econ) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_unemployment(unemployment_rate),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "Unemployment: ", round(unemployment_rate, 1), "%"
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_unemployment,
    values = ~unemployment_rate,
    title = "Unemployment (%)",
    opacity = 0.7,
    na.label = "No data"
  )
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Households Without Vehicle"

# Create color palette
pal_vehicle <- colorNumeric(
  palette = "Reds",
  domain = cu_data_econ$pct_no_vehicle[!is.na(cu_data_econ$pct_no_vehicle)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_econ) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_vehicle(pct_no_vehicle),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "No Vehicle: ", round(pct_no_vehicle, 1), "%"
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_vehicle,
    values = ~pct_no_vehicle,
    title = "No Vehicle (%)",
    opacity = 0.7,
    na.label = "No data"
  )
```

## Row {height="35%"}

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Educational Attainment Distribution"

plot_ly(
  x = ~cu_data_econ$pct_bachelor_plus,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#9e9ac8", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Bachelor's Degree or Higher (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 50, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Unemployment Rate Distribution"

plot_ly(
  x = ~cu_data_econ$unemployment_rate,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#fd8d3c", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Unemployment Rate (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 50, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "No Vehicle Distribution"

plot_ly(
  x = ~cu_data_econ$pct_no_vehicle,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#fc4e2a", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Households Without Vehicle (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 50, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {height="10%"}

### Column
```{r}
#| echo: false

div(
  style = "padding: 10px; background-color: #f8f9fa; border-left: 4px solid #6c757d; margin: 5px; font-size: 13px;",
  tags$p(
    style = "margin: 0; color: #495057;",
    tags$strong("Source: "),
    "U.S. Census Bureau, American Community Survey 5-Year Estimates (2018-2022). Retrieved via tidycensus."
  )
)
```

# Life Expectancy

## Row {height="55%"}

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Life Expectancy by Census Tract (2010-2015)"

# Read the mortality data
cu_mortality <- read.csv("/Users/jakecox/rstudio/cu-geo/cu-mortality-clean.csv")

# Join the data
cu_data_life <- cu_geo %>%
  left_join(cu_mortality, by = "NAMELSAD")

# Create color palette
pal_life_exp <- colorNumeric(
  palette = "RdYlGn",
  domain = cu_data_life$Life_Expectancy_2010_2015[!is.na(cu_data_life$Life_Expectancy_2010_2015)],
  na.color = "grey"
)

# Create leaflet map
leaflet(cu_data_life) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal_life_exp(Life_Expectancy_2010_2015),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0(
      "<strong>", NAMELSAD, "</strong><br>",
      "Life Expectancy: ", 
      ifelse(is.na(Life_Expectancy_2010_2015), 
             "Data Suppressed", 
             paste0(round(Life_Expectancy_2010_2015, 1), " years")), "<br>",
      "Population: ", format(Population, big.mark = ",")
    ) %>% lapply(HTML)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_life_exp,
    values = ~Life_Expectancy_2010_2015,
    title = "Life Expectancy<br>(years)",
    opacity = 0.7,
    na.label = "No data"
  )
```

## Row {height="35%"}

### Column {width="50%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Life Expectancy Distribution"

plot_ly(
  x = ~cu_data_life$Life_Expectancy_2010_2015,
  type = "histogram",
  nbinsx = 15,
  marker = list(color = "#74c476", line = list(color = "white", width = 1))
) %>%
  layout(
    xaxis = list(title = "Life Expectancy (years)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Census Tracts", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 55, r = 20, t = 30, b = 50)
  ) %>%
  config(displayModeBar = FALSE)
```

### Column {width="50%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Life Expectancy Gap Analysis"

# Calculate life expectancy categories
cu_data_summary <- cu_data_life %>%
  st_drop_geometry() %>%
  filter(!is.na(Life_Expectancy_2010_2015)) %>%
  mutate(
    LE_Category = case_when(
      Life_Expectancy_2010_2015 < 75 ~ "Below 75",
      Life_Expectancy_2010_2015 >= 75 & Life_Expectancy_2010_2015 < 80 ~ "75-80",
      Life_Expectancy_2010_2015 >= 80 & Life_Expectancy_2010_2015 < 85 ~ "80-85",
      Life_Expectancy_2010_2015 >= 85 ~ "85+"
    ),
    LE_Category = factor(LE_Category, levels = c("Below 75", "75-80", "80-85", "85+"))
  )

category_counts <- cu_data_summary %>%
  group_by(LE_Category) %>%
  summarise(
    Count = n(),
    Avg_LE = mean(Life_Expectancy_2010_2015, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(LE_Category)

# Calculate the gap
le_min <- min(cu_data_life$Life_Expectancy_2010_2015, na.rm = TRUE)
le_max <- max(cu_data_life$Life_Expectancy_2010_2015, na.rm = TRUE)
le_gap <- round(le_max - le_min, 1)

# Create bar chart
plot_ly(
  data = category_counts,
  x = ~LE_Category,
  y = ~Count,
  type = 'bar',
  marker = list(
    color = c('#d7191c', '#fdae61', '#a6d96a', '#1a9641'),
    line = list(color = 'white', width = 1)
  ),
  text = ~Count,
  textposition = 'inside',
  textfont = list(color = 'white', size = 12),
  hoverinfo = 'text',
  hovertext = ~paste0(Count, " tracts<br>Avg: ", round(Avg_LE, 1), " yrs")
) %>%
  layout(
    xaxis = list(
      title = "Life Expectancy (years)", 
      gridcolor = "#f0f0f0",
      tickfont = list(size = 11)
    ),
    yaxis = list(
      title = "Census Tracts", 
      gridcolor = "#f0f0f0",
      range = c(0, max(category_counts$Count) + 2)
    ),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(t = 45, b = 50, l = 55, r = 20),
    annotations = list(
      list(
        x = 0.5,
        y = 1.06,
        xref = "paper",
        yref = "paper",
        text = paste0("<b>", le_gap, " year gap</b> between tracts"),
        showarrow = FALSE,
        font = list(size = 11, color = "#d7191c"),
        xanchor = "center"
      )
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {height="10%"}

### Column
```{r}
#| echo: false

div(
  style = "padding: 10px; background-color: #f8f9fa; border-left: 4px solid #6c757d; margin: 5px; font-size: 13px;",
  tags$p(
    style = "margin: 0; color: #495057;",
    tags$strong("Data Note: "),
    "Some tracts have suppressed data due to small populations. ",
    tags$strong("Source: "),
    "Champaign-Urbana Public Health District, 2010-2015."
  )
)
```

# Bivariate Analysis

## Row {height="45%"}

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Life Expectancy vs. Poverty Rate"

# Read fresh geo data and convert GEOID immediately
cu_geo_bv <- st_read("/Users/jakecox/rstudio/cu-geo/cu-tracts-fixed.json", quiet = TRUE)
cu_geo_bv$GEOID <- as.character(cu_geo_bv$GEOID)

# Read CSV data files
cu_poverty_bv <- read.csv("/Users/jakecox/rstudio/cu-geo/cu-poverty_revised.csv")
cu_sdoh_bv <- read.csv("/Users/jakecox/rstudio/cu-geo/cu-sdoh.csv")
cu_mortality_bv <- read.csv("/Users/jakecox/rstudio/cu-geo/cu-mortality-clean.csv")

# Get ACS education data
education_bivar <- get_acs(
  geography = "tract", state = "IL", county = "Champaign",
  variables = c(
    bachelors    = "B15003_022",
    masters      = "B15003_023",
    professional = "B15003_024",
    doctorate    = "B15003_025",
    total_pop    = "B15003_001"
  ),
  year = 2022, survey = "acs5", output = "wide"
)

# Process education data using pure base R
education_bivar$pct_bachelor_plus <- ((education_bivar$bachelorsE + education_bivar$mastersE + 
                                        education_bivar$professionalE + education_bivar$doctorateE) / 
                                        education_bivar$total_popE) * 100
education_bivar$GEOID <- as.character(education_bivar$GEOID)
education_bivar <- data.frame(GEOID = education_bivar$GEOID, 
                               pct_bachelor_plus = education_bivar$pct_bachelor_plus)

# Create base dataset
cu_bivar_base <- st_drop_geometry(cu_geo_bv)

# Join using base R merge to avoid dplyr type issues
cu_bivar <- merge(cu_bivar_base, cu_poverty_bv, by = "NAMELSAD", all.x = TRUE)
cu_bivar <- merge(cu_bivar, cu_sdoh_bv, by = "NAMELSAD", all.x = TRUE)
cu_bivar <- merge(cu_bivar, cu_mortality_bv, by = "NAMELSAD", all.x = TRUE)
cu_bivar <- merge(cu_bivar, education_bivar, by = "GEOID", all.x = TRUE)
cu_bivar <- cu_bivar[!is.na(cu_bivar$Life_Expectancy_2010_2015), ]

# Calculate correlation
cor_pov <- cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$Poverty.Rate, use = "complete.obs")

# Scatter plot
plot_ly(
  data = cu_bivar,
  x = ~Poverty.Rate,
  y = ~Life_Expectancy_2010_2015,
  type = 'scatter',
  mode = 'markers',
  marker = list(
    color = '#d7191c',
    size = 10,
    opacity = 0.7,
    line = list(color = 'white', width = 1)
  ),
  text = ~paste0("<b>", NAMELSAD, "</b><br>",
                 "Poverty: ", round(Poverty.Rate, 1), "%<br>",
                 "Life Exp: ", round(Life_Expectancy_2010_2015, 1), " yrs"),
  hoverinfo = 'text'
) %>%
  layout(
    xaxis = list(title = "Poverty Rate (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Life Expectancy (years)", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 55, r = 20, t = 40, b = 50),
    annotations = list(
      list(
        x = 0.98, y = 0.98,
        xref = "paper", yref = "paper",
        text = paste0("r = ", round(cor_pov, 2)),
        showarrow = FALSE,
        font = list(size = 14, color = "#d7191c"),
        xanchor = "right"
      )
    )
  ) %>%
  config(displayModeBar = FALSE)
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Life Expectancy vs. Median Income"

# Calculate correlation
cor_inc <- cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$median_income, use = "complete.obs")

# Scatter plot
plot_ly(
  data = cu_bivar,
  x = ~median_income,
  y = ~Life_Expectancy_2010_2015,
  type = 'scatter',
  mode = 'markers',
  marker = list(
    color = '#1a9641',
    size = 10,
    opacity = 0.7,
    line = list(color = 'white', width = 1)
  ),
  text = ~paste0("<b>", NAMELSAD, "</b><br>",
                 "Income: $", format(median_income, big.mark = ","), "<br>",
                 "Life Exp: ", round(Life_Expectancy_2010_2015, 1), " yrs"),
  hoverinfo = 'text'
) %>%
  layout(
    xaxis = list(title = "Median Household Income ($)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Life Expectancy (years)", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 55, r = 20, t = 40, b = 50),
    annotations = list(
      list(
        x = 0.98, y = 0.98,
        xref = "paper", yref = "paper",
        text = paste0("r = ", round(cor_inc, 2)),
        showarrow = FALSE,
        font = list(size = 14, color = "#1a9641"),
        xanchor = "right"
      )
    )
  ) %>%
  config(displayModeBar = FALSE)
```

### Column
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Life Expectancy vs. Education"

# Calculate correlation
cor_edu <- cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$pct_bachelor_plus, use = "complete.obs")

# Scatter plot
plot_ly(
  data = cu_bivar,
  x = ~pct_bachelor_plus,
  y = ~Life_Expectancy_2010_2015,
  type = 'scatter',
  mode = 'markers',
  marker = list(
    color = '#756bb1',
    size = 10,
    opacity = 0.7,
    line = list(color = 'white', width = 1)
  ),
  text = ~paste0("<b>", NAMELSAD, "</b><br>",
                 "Bachelor's+: ", round(pct_bachelor_plus, 1), "%<br>",
                 "Life Exp: ", round(Life_Expectancy_2010_2015, 1), " yrs"),
  hoverinfo = 'text'
) %>%
  layout(
    xaxis = list(title = "Bachelor's Degree or Higher (%)", gridcolor = "#f0f0f0"),
    yaxis = list(title = "Life Expectancy (years)", gridcolor = "#f0f0f0"),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 55, r = 20, t = 40, b = 50),
    annotations = list(
      list(
        x = 0.98, y = 0.98,
        xref = "paper", yref = "paper",
        text = paste0("r = ", round(cor_edu, 2)),
        showarrow = FALSE,
        font = list(size = 14, color = "#756bb1"),
        xanchor = "right"
      )
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {height="45%"}

### Column {width="60%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Correlation Matrix"

# Create correlation matrix using base R
cor_vars <- cu_bivar[, c("Life_Expectancy_2010_2015", "Poverty.Rate", 
                          "median_income", "insurance_coverage", "pct_bachelor_plus")]
names(cor_vars) <- c("Life Expectancy", "Poverty Rate", "Median Income", 
                      "Insurance Coverage", "Bachelor's+")

cor_matrix <- cor(cor_vars, use = "complete.obs")

# Create heatmap
plot_ly(
  x = colnames(cor_matrix),
  y = rownames(cor_matrix),
  z = cor_matrix,
  type = "heatmap",
  colorscale = list(
    list(0, "#d7191c"),
    list(0.5, "#ffffbf"),
    list(1, "#1a9641")
  ),
  zmin = -1,
  zmax = 1,
  text = round(cor_matrix, 2),
  texttemplate = "%{text}",
  textfont = list(size = 12, color = "black"),
  hoverinfo = "z",
  colorbar = list(
    title = list(text = "r", font = list(size = 12)),
    tickvals = c(-1, -0.5, 0, 0.5, 1),
    ticktext = c("-1", "-0.5", "0", "0.5", "1"),
    len = 0.75,
    thickness = 15
  )
) %>%
  layout(
    xaxis = list(title = "", tickangle = -45, tickfont = list(size = 11)),
    yaxis = list(title = "", tickfont = list(size = 11), autorange = "reversed"),
    plot_bgcolor = "white",
    paper_bgcolor = "white",
    margin = list(l = 100, r = 20, t = 30, b = 100)
  ) %>%
  config(displayModeBar = FALSE)
```

### Column {width="40%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| title: "Key Correlations with Life Expectancy"

# Calculate correlations with life expectancy using base R
cor_data <- data.frame(
  Variable = c("Median Income", "Bachelor's+", "Insurance", "Poverty Rate"),
  Correlation = c(
    cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$median_income, use = "complete.obs"),
    cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$pct_bachelor_plus, use = "complete.obs"),
    cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$insurance_coverage, use = "complete.obs"),
    cor(cu_bivar$Life_Expectancy_2010_2015, cu_bivar$Poverty.Rate, use = "complete.obs")
  )
)
cor_data$Variable <- factor(cor_data$Variable, levels = cor_data$Variable[order(cor_data$Correlation)])
cor_data$Color <- ifelse(cor_data$Correlation > 0, "#1a9641", "#d7191c")

plot_ly(
  data = cor_data,
  x = ~Correlation,
  y = ~Variable,
  type = 'bar',
  orientation = 'h',
  marker = list(
    color = ~Color,
    line = list(color = 'white', width = 1)
  ),
  text = ~paste0(round(Correlation, 2)),
  textposition = 'outside',
  textfont = list(size = 12),
  hoverinfo = 'text',
  hovertext = ~paste0(Variable, ": r = ", round(Correlation, 2))
) %>%
  layout(
    xaxis = list(
      title = "Correlation Coefficient (r)",
      gridcolor = "#f0f0f0",
      range = c(-1, 1),
      zeroline = TRUE,
      zerolinecolor = "#666",
      zerolinewidth = 1
    ),
    yaxis = list(title = "", tickfont = list(size = 11)),
    plot_bgcolor = "#fafafa",
    paper_bgcolor = "white",
    margin = list(l = 100, r = 50, t = 30, b = 50),
    showlegend = FALSE
  ) %>%
  config(displayModeBar = FALSE)
```

## Row {height="10%"}

### Column
```{r}
#| echo: false

div(
  style = "padding: 10px; background-color: #f8f9fa; border-left: 4px solid #6c757d; margin: 5px; font-size: 13px;",
  tags$p(
    style = "margin: 0; color: #495057;",
    tags$strong("Note: "),
    "Pearson correlation coefficients (r) range from -1 to 1. Values closer to ±1 indicate stronger relationships. ",
    tags$strong("Sources: "),
    "ACS 2018-2022; CUPHD 2010-2015."
  )
)
```
